---
title: "Marker List Analysis"
author: "Brent Biddy"
output:
    html_document:
        code_folding: "show"
        toc: true
        toc_float: true
        fig_width: 10
        fig_height: 10
params:
    marker_path: "./marker_list.yaml"
    rds_path: "./seurat_object.RDS"
    annotation_path: "./annotations.csv"
    cluster_name: "xenium_clusters"
    coi: !expr c(19)
    roi: 3
    filt_umap_1: 0
---

# Introduction

This analysis examines spatial transcriptomics data from human ovary to identify and characterize different cell types using marker gene expression patterns. We analyze Xenium spatial data from ROI `r knitr::combine_words(params$roi)`, focusing on how cell type markers correspond to the clustering structure observed in the data.
This analysis will focus mainly on identifying oocytes.

The main objectives are:

1. **Load and validate marker gene lists** for oocytes, granulosa cells, theca cells, upC1, and upC2 cell types
2. **Calculate cell type scores** based on average expression of the marker gene list
3. **Visualize expression patterns** across clusters using heatmaps, UMAPs, and violin plots
4. **Perform focused sub-clustering** on specific clusters of interest to identify potential subpopulations of oocytes

# Setup and Configuration

```{r setup}
# Set global chunk options to suppress warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = TRUE) 

# Load required libraries
library(Seurat)        # Single-cell analysis
library(yaml)          # YAML file parsing
library(dplyr)         # Data manipulation
library(stringr)       # String operations
library(cowplot)       # Plot arrangements
library(ggplot2)       # Visualization
# library(VennDiagram)   # Venn diagrams

# Set plotting options
options(repr.plot.width = 12, repr.plot.height = 7.5, repr.plot.res = 200)
options(warn = -1)  # Suppress warnings

if(params$cluster_name == "seurat_clusters"){
    umap_use <- "umap"
} else{
    umap_use <- "xenium_umap"
}
```

# Data Loading and Preprocessing

## Load Marker List

We begin by loading a curated list of marker genes for each cell type. These markers will serve as the basis for calculating cell type enrichment scores.

```{r load-markers}
# Display the marker file path for verification
print(params$marker_path)

# Read YAML file containing cell type markers
marker_list <- yaml::read_yaml(file = params$marker_path)

# Extract marker genes for each cell type
marker_list <- lapply(marker_list$cell_types, '[[', "markers")

# Flatten nested lists to single character vectors
marker_list <- lapply(marker_list, unlist)

# Convert all gene names to uppercase for consistency
marker_list <- lapply(marker_list, str_to_upper)

# Create a data frame from the marker list for easier manipulation
marker_dat <- data.frame(unlist(marker_list))
colnames(marker_dat)[1] <- "Gene"

# Extract cell type from rownames (format: celltype.number)
marker_dat[["cell_type"]] <- rownames(marker_dat)
marker_dat[["cell_type"]] <- strsplit(marker_dat[["cell_type"]], split = ".", fixed = TRUE)
marker_dat[["cell_type"]] <- unlist(lapply(marker_dat[["cell_type"]], "[[", 1))
```

## Load Stage Annotations

We load the follicle developmental stage annotations that Taylor provided.

```{r load-annotations}
# Read follicle stage annotation data
stage_dat <- data.table::fread(params$annotation_path, data.table = FALSE)

# Filter and prepare stage data for ROI3
stage_dat |>
    filter(.data[["Donor ROI"]] == paste0("ROI", params$roi)) |>
    select(c("Donor ROI", "Image", "Stage Labels", "Quality")) |>
    mutate(cell = Image) ->
    stage_dat

# Extract cell ID from image filename
# Format: [prefix]_[cell_id].png
stage_dat[["cell"]] <- strsplit(x = stage_dat[["cell"]], split = "_")
stage_dat[["cell"]] <- sapply(stage_dat[["cell"]], "[[", 2)
stage_dat[["cell"]] <- gsub(pattern = ".png", replacement = "", x = stage_dat[["cell"]])
```

## Load Seurat Object

The Seurat object contains the preprocessed Xenium spatial transcriptomics data, including normalized expression values, dimensionality reductions, and cluster assignments.

```{r load-seurat}
# Load the pre-processed Seurat object containing spatial transcriptomics data
seurat <- readRDS(params$rds_path)
seurat
```

## Merge Annotations with Seurat Object

By integrating stage annotations into the Seurat object metadata, we can overlay Taylor's follicle annotations onto our clustering results and examine whether clusters correspond to known developmental stages.

```{r merge-annotations}
# Join stage and quality annotations to the Seurat metadata
# This allows us to analyze clusters by developmental stage
seurat[[]] <- left_join(
    seurat[[]], 
    stage_dat[c("cell", "Stage Labels", "Quality")], 
    by = join_by(cell_id == cell)
)
```

# Marker Gene Analysis

## Marker Summary Statistics

The number of markers for each cell type.

```{r marker-summary}
# Count the number of markers per cell type
marker_dat |>
    select(cell_type, Gene) |>
    group_by(cell_type) |>
    summarise(Markers = n())
```

## Identify Duplicate Markers

Some genes may serve as markers for multiple cell types. Identifying these helps us understand which markers are cell type-specific versus broadly expressed. Genes that appear in multiple marker lists are labeled as "multi" to prevent misinterpretation.

```{r duplicate-markers}
# Find genes that are markers for multiple cell types
# This helps identify genes with broader expression patterns
duplicated_markers <- marker_dat[
    duplicated(x = marker_dat[["Gene"]]) | 
    duplicated(x = marker_dat[["Gene"]], fromLast = TRUE), 
] |> 
    arrange(Gene) |> 
    tibble::remove_rownames()

duplicated_markers
```

## Label Multi-Cell Type Markers

```{r label-multi-markers}
# Assign "multi" label to genes that appear in multiple cell types
# This prevents them from being associated with a single cell type
marker_dat[marker_dat[["Gene"]] %in% marker_dat[["Gene"]][duplicated(marker_dat[["Gene"]])], "cell_type"] <- "multi"
```

# Color Scheme Setup

Consistent color schemes across all visualizations make it easier to track clusters and cell types throughout the analysis.

```{r setup-colors}
# Generate color palette for clusters
clusters <- sort(unique(seurat[[]][[params$cluster_name]]), na.last = TRUE)
cluster_cols <- pals::brewer.bugn(n = length(clusters))
names(cluster_cols) <- clusters

# Remove NA cluster colors for heatmap (if any)
heatmap_cluster_cols <- cluster_cols[!is.na(names(cluster_cols))]

# Generate color palette for cell types
cell_types <- sort(unique(marker_dat$cell_type), na.last = TRUE)
celltype_cols <- pals::stepped(n = length(cell_types))
names(celltype_cols) <- cell_types
```

## Cluster Distribution

Before diving into marker expression, it's useful to see how cells are distributed across clusters. This gives us a sense of cluster sizes and helps identify rare versus common cell populations.

```{r cluster-distribution}
# Display the number of cells in each cluster
table(seurat[[]][[params$cluster_name]])
```

## Calculate Average Expression Scores

For each cell type, we calculate a composite score by averaging the expression of all its marker genes. This approach provides a single metric that summarizes how well each cell matches each cell type's marker gene profile. We also calculate a differential score between upC1 and upC2 to identify cells that are more similar to one or the other.

```{r calculate-avg-expression}
# For each cell type, calculate the average expression of its marker genes
# This creates a composite "cell type score" for each cell
for(ct in unique(marker_dat$cell_type)){
    # Get marker genes for this cell type
    marker_dat |>
        filter(cell_type == ct) |>
        pull(Gene) -> gene_list
    
    # Fetch expression data for these markers
    exp_dat <- FetchData(object = seurat, vars = gene_list)
    
    # Calculate mean expression across all markers
    exp_dat$avg_exp <- rowMeans(exp_dat)
    
    # Add as metadata to Seurat object
    seurat[[paste0(ct, "_avg")]] <- exp_dat$avg_exp
}

# Calculate differential score between upC1 and upC2
# Positive values indicate upC1-like, negative indicate upC2-like
seurat[["upC_diff"]] <- seurat[["upC1_avg"]] - seurat[["upC2_avg"]]
```

# Heatmap Visualization

## Overview

The heatmap provides a comprehensive view of marker gene expression across all clusters. Rows are organized by cell type (indicated by the left annotation), while columns are grouped by cluster identity (top annotation). This visualization helps us identify:

- Which clusters express markers for specific cell types
- Whether marker expression patterns align with cluster boundaries
- Potential mixed or transitional populations

**How to interpret**: Dark purple indicates high expression, yellow indicates low expression.

## Prepare Heatmap Data

```{r prepare-heatmap-data}
# Generate heatmap data using Seurat's DoHeatmap function
# Downsample to 200 cells per cluster for visualization
heat_dat <- DoHeatmap(
    object = subset(seurat, downsample = 200),  
    group.by = params$cluster_name, 
    features = unlist(marker_list), 
    label = FALSE
)$data |> 
    tidyr::drop_na()

# Convert to wide format matrix for ComplexHeatmap
heat_dat |>
    select(Feature, Cell, Expression) |>
    tidyr::pivot_wider(names_from = Cell, values_from = Expression, values_fill = NA) |>
    tibble::column_to_rownames("Feature") |>
    as.matrix() -> heat_mtx

# Create column annotations (cluster identities)
heat_dat |>
    select(Cell, Identity) |>
    distinct(Cell, .keep_all = TRUE) |>
    mutate(Identity = as.factor(Identity), Cell = as.vector(Cell)) |>
    filter(Cell %in% colnames(heat_mtx)) |>
    arrange(Identity) |>
    tibble::column_to_rownames("Cell") ->
    colAnnot

# Create row annotations (cell type associations)
heat_dat |>
    left_join(y = marker_dat, by = join_by(Feature == Gene)) |>
    select(Feature, cell_type) |>
    filter(Feature %in% rownames(heat_mtx)) |>
    distinct(Feature, .keep_all = TRUE) ->
    rowAnnot

# Reorder matrix columns to match annotation order
heat_mtx <- heat_mtx[, rownames(colAnnot)]
```

## Generate Complex Heatmap

```{r generate-heatmap, fig.width=12, fig.height=10}
# Create comprehensive heatmap with annotations
ComplexHeatmap::Heatmap(
    matrix = heat_mtx,
    row_names_side = "left",
    col = viridis::viridis(n = 100),  # Use viridis color scale
    show_column_names = FALSE,         # Hide cell names (too many)
    show_row_names = TRUE,             # Show gene names
    show_column_dend = FALSE,          # No column dendrogram
    show_row_dend = FALSE,             # No row dendrogram
    cluster_columns = FALSE,           # Don't cluster columns
    cluster_rows = FALSE,              # Don't cluster rows
    cluster_column_slices = FALSE,     # Don't cluster column groups
    cluster_row_slices = FALSE,        # Don't cluster row groups
    column_split = colAnnot$Identity,  # Split by cluster
    row_split = rowAnnot$cell_type,    # Split by cell type
    # Top annotation shows cluster identity
    top_annotation = ComplexHeatmap::HeatmapAnnotation(
        Cluster = colAnnot$Identity, 
        col = list("Cluster" = heatmap_cluster_cols), 
        annotation_name_gp = gpar(fontsize = 10), 
        show_legend = FALSE
    ),
    # Left annotation shows cell type
    left_annotation = ComplexHeatmap::rowAnnotation(
        CellType = rowAnnot$cell_type, 
        col = list("CellType" = celltype_cols), 
        annotation_name_gp = gpar(fontsize = 10), 
        show_legend = FALSE
    ),
    row_names_gp = gpar(fontsize = 6),
    heatmap_legend_param = list(
        title = "Counts",
        legend_height = unit(6, "cm"),
        title_position = "leftcenter-rot"
    ),
    show_parent_dend_line = FALSE
)
```

# Cell Type Expression Patterns

In this section, we examine the spatial and cluster-level distribution of each cell type signature. For each cell type, we show:

1. **UMAP (left)**: Cells colored by their cell type score
2. **Cluster UMAP (middle)**: The original cluster assignments for reference
3. **Violin plot (right)**: Distribution of cell type scores across clusters

These visualizations help identify which clusters are enriched for specific cell types and whether cell type identities align with the clustering structure.

## Oocyte Markers


```{r oocyte-plots, fig.width=12, fig.height=8}
# Create cluster UMAP for reference
cluster_umap <- DimPlot(
    object = seurat, 
    group.by = params$cluster_name, 
    reduction = umap_use, 
    cols = cluster_cols, 
    label = TRUE, 
    repel = TRUE
) + 
    theme(legend.position = "right", legend.key.size = unit(2, "pt"))

# UMAP colored by average oocyte marker expression
oocyte_umap <- FeaturePlot(
    object = seurat, 
    features = "oocyte_avg", 
    reduction = umap_use
) + 
    theme_bw() + 
    theme(legend.position = "right")

# Violin plot of oocyte score by cluster
oocyte_vln <- VlnPlot(
    object = seurat, 
    features = "oocyte_avg", 
    pt.size = 0, 
    group.by = params$cluster_name
) + 
    scale_fill_manual(values = cluster_cols) + 
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) + 
    theme_linedraw() + 
    theme(legend.position = "none")

# Combine plots
plot_grid(
    oocyte_umap,
    cluster_umap,
    oocyte_vln,
    align = 'vh'
)
```

## Theca Cell Markers

```{r theca-plots, fig.width=12, fig.height=8}
# UMAP colored by average theca marker expression
theca_umap <- FeaturePlot(
    object = seurat, 
    features = "theca_cell_avg", 
    min.cutoff = 0.0, 
    reduction = umap_use
) + 
    theme_bw() + 
    theme(legend.position = "right")

# Violin plot of theca score by cluster
theca_vln <- VlnPlot(
    object = seurat, 
    features = "theca_cell_avg", 
    pt.size = 0, 
    group.by = params$cluster_name
) + 
    scale_fill_manual(values = cluster_cols) + 
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) + 
    theme_linedraw() + 
    theme(legend.position = "none")

# Combine plots
plot_grid(
    theca_umap,
    cluster_umap,
    theca_vln,
    align = 'vh'
)
```

## Granulosa Cell Markers

```{r granulosa-plots, fig.width=12, fig.height=8}
# UMAP colored by average granulosa marker expression
granulosa_umap <- FeaturePlot(
    object = seurat, 
    features = "granulosa_cell_avg", 
    min.cutoff = 0.0, 
    reduction = umap_use
) + 
    theme_bw() + 
    theme(legend.position = "right")

# Violin plot of granulosa score by cluster
granulosa_vln <- VlnPlot(
    object = seurat, 
    features = "granulosa_cell_avg", 
    pt.size = 0, 
    group.by = params$cluster_name
) + 
    scale_fill_manual(values = cluster_cols) + 
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) + 
    theme_linedraw() + 
    theme(legend.position = "none")

# Combine plots
plot_grid(
    granulosa_umap,
    cluster_umap,
    granulosa_vln,
    align = 'vh'
)
```

## UPC1 Markers

```{r upc1-plots, fig.width=12, fig.height=8}
# UMAP colored by average UPC1 marker expression
upc1_umap <- FeaturePlot(
    object = seurat, 
    features = "upC1_avg", 
    min.cutoff = 1.0, 
    reduction = umap_use
) + 
    theme_bw() + 
    theme(legend.position = "right")

# Violin plot of UPC1 score by cluster
upc1_vln <- VlnPlot(
    object = seurat, 
    features = "upC1_avg", 
    pt.size = 0, 
    group.by = params$cluster_name
) + 
    scale_fill_manual(values = cluster_cols) + 
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) + 
    theme_linedraw() + 
    theme(legend.position = "none")

# Combine plots
plot_grid(
    upc1_umap,
    cluster_umap,
    upc1_vln,
    align = 'vh'
)
```

## UPC2 Markers

```{r upc2-plots, fig.width=12, fig.height=8}
# UMAP colored by average UPC2 marker expression
upc2_umap <- FeaturePlot(
    object = seurat, 
    features = "upC2_avg", 
    min.cutoff = 0.0, 
    reduction = umap_use
) + 
    theme_bw() + 
    theme(legend.position = "right")

# Violin plot of UPC2 score by cluster
upc2_vln <- VlnPlot(
    object = seurat, 
    features = "upC2_avg", 
    pt.size = 0, 
    group.by = params$cluster_name
) + 
    scale_fill_manual(values = cluster_cols) + 
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) + 
    theme_linedraw() + 
    theme(legend.position = "none")

# Combine plots
plot_grid(
    upc2_umap,
    cluster_umap,
    upc2_vln,
    align = 'vh'
)
```

## UPC1 vs UPC2 Differential Score

**Interpretation**: The differential score (UPC1 - UPC2) highlights cells along a spectrum:
- **Positive values (red)**: More UPC1-like
- **Negative values (blue)**: More UPC2-like
- **Values near zero**: Mixed or intermediate states

This can reveal whether UPC1 and UPC2 represent discrete populations or ends of a continuous spectrum.

```{r upc-diff-plots, fig.width=12, fig.height=8}
# UMAP showing UPC1-UPC2 difference
# Positive values = more UPC1-like, negative = more UPC2-like
upc_umap <- FeaturePlot(
    object = seurat, 
    features = "upC_diff", 
    min.cutoff = 0.0, 
    reduction = umap_use
) + 
    theme_bw() + 
    theme(legend.position = "right")

# Violin plot of differential score by cluster
upc_vln <- VlnPlot(
    object = seurat, 
    features = "upC_diff", 
    pt.size = 0, 
    group.by = params$cluster_name
) + 
    scale_fill_manual(values = cluster_cols) + 
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) + 
    theme_linedraw() + 
    theme(legend.position = "none")

# Combine plots
plot_grid(
    upc_umap,
    cluster_umap,
    upc_vln,
    align = 'vh'
)
```

## Combined Violin Plots

This stacked view allows for direct comparison of cell type score distributions across all clusters.

# Stack all cell type violin plots for easy comparison
```{r combined-violins, fig.width=10, fig.height=12}
plot_grid(
    oocyte_vln + theme(legend.position = "none") + xlab(""),
    granulosa_vln + theme(legend.position = "none") + xlab(""),
    theca_vln + theme(legend.position = "none") + xlab(""),
    upc1_vln + theme(legend.position = "none") + xlab(""),
    upc2_vln + theme(legend.position = "none"),
    ncol = 1,
    align = "vh"
)
```

# Focused Cluster Analysis

## Motivation

Sometimes individual clusters contain heterogeneity that isn't captured by the initial clustering. By selecting a cluster of interest and re-analyzing it in isolation, we can:

- Identify subpopulations within the cluster
- Examine whether marker-based dimensionality reduction reveals different structure

## Select Cluster of Interest

```{r select-cluster}
# Define which cluster to analyze in depth
# cluster_of_interest <- as.character(params$coi)

# Total number of unique marker genes
length(unique(marker_dat$Gene))
```

## Subset and Re-analyze Cluster

We extract cluster `r knitr::combine_words(params$coi)` and perform a complete re-analysis, including both standard (all genes) and marker-based (curated markers only) dimensionality reduction. This allows us to compare whether using only biologically-relevant markers reveals different clustering structure.

```{r subset-cluster}
# Store original cluster assignments
seurat[["seurat_clusters_full"]] <- seurat[["seurat_clusters"]]

# Subset to only cells in cluster of interest
# cluster_seurat <- subset(seurat, seurat_clusters %in% params$coi)
cluster_seurat <- subset(seurat, !!sym(params$cluster_name) %in% params$coi)
cluster_seurat
```

## Re-process Subset

```{r reprocess-cluster}
# Perform standard Seurat workflow on the subset
# This allows us to find sub-structure within the cluster

# Normalize counts
cluster_seurat <- NormalizeData(object = cluster_seurat)

# Find highly variable genes
cluster_seurat <- FindVariableFeatures(object = cluster_seurat)

# Scale data
cluster_seurat <- ScaleData(object = cluster_seurat)

# Run PCA on all genes
cluster_seurat <- RunPCA(object = cluster_seurat)

# Run PCA using only marker genes
cluster_seurat <- RunPCA(
    object = cluster_seurat, 
    features = unique(marker_dat$Gene), 
    reduction.name = "pca_markers"
)

# Find neighbors using standard PCA
cluster_seurat <- FindNeighbors(object = cluster_seurat, dims = 1:30)

# Find neighbors using marker-based PCA
cluster_seurat <- FindNeighbors(
    object = cluster_seurat, 
    dims = 1:30, 
    reduction = "pca_markers", 
    graph.name = "marker_graph"
)

# Cluster using standard graph
cluster_seurat <- FindClusters(object = cluster_seurat)

# Cluster using marker-based graph
cluster_seurat <- FindClusters(
    object = cluster_seurat, 
    graph.name = "marker_graph", 
    cluster.name = "marker_clusters"
)

# Generate UMAP from standard PCA
cluster_seurat <- RunUMAP(object = cluster_seurat, dims = 1:30)

# Generate UMAP from marker-based PCA
cluster_seurat <- RunUMAP(
    object = cluster_seurat, 
    dims = 1:30, 
    reduction = "pca_markers", 
    reduction.name = "marker_umap"
)
```

# Spatial Visualization

Spatial plots show where cells are located in the original tissue context. This is important for understanding whether identified clusters correspond to anatomically distinct regions or are spatially intermingled.
This comparison shows the original cluster assignments (left) alongside just the cells from cluster `r knitr::combine_words(params$coi)` (right).

## Tissue Distribution - Cluster of Interest

```{r tissue-cluster-highlight, fig.width=12, fig.height=6}
# Plot spatial distribution of all clusters
SeuratObject::GetTissueCoordinates(object = seurat) |>
    left_join(seurat[[]], by = join_by(cell == cell_id)) |>
    ggplot(aes(x = x, y = y, color = as.factor(!!sym(params$cluster_name)))) + 
    geom_point(size = 0) ->
    plt_one

# Highlight only the cluster of interest
SeuratObject::GetTissueCoordinates(object = seurat) |>
    left_join(seurat[[]], by = join_by(cell == cell_id)) |>
    mutate(coi = if_else(!!sym(params$cluster_name) %in% params$coi, !!sym(params$cluster_name), NA)) |>
    mutate(coi = as.factor(coi)) |>
    ggplot(aes(x = x, y = y, color = as.factor(coi))) + 
    geom_point(size = 0) ->
    plt_two

# Display side by side
plot_grid(
    plt_one + coord_flip(),
    plt_two + coord_flip()
)
```

## Tissue Distribution - Original vs Subset

This comparison shows the original cluster assignments (left) alongside just the new subset cluster ids (right).

```{r tissue-original-vs-subset, fig.width=12, fig.height=6}
# Original cluster assignments in tissue space
SeuratObject::GetTissueCoordinates(object = seurat) |>
    left_join(seurat[[]], by = join_by(cell == cell_id)) |>
    ggplot(aes(x = x, y = y, color = as.factor(!!sym(params$cluster_name)))) + 
    geom_point(size = 0) ->
    plt_one

# Subset cluster assignments in tissue space
SeuratObject::GetTissueCoordinates(object = seurat) |>
    left_join(cluster_seurat[[]], by = join_by(cell == cell_id)) |>
    ggplot(aes(x = x, y = y, color = seurat_clusters)) + 
    geom_point(size = 0) ->
    plt_two

# Display side by side
plot_grid(
    plt_one + coord_flip(),
    plt_two + coord_flip()
)
```

## Tissue Distribution - Marker-based Clustering

**Key question**: Do the marker-based sub-clusters (right panel) show spatial organization? If so, this suggests biologically meaningful heterogeneity within cluster `r knitr::combine_words(params$coi)`.

```{r tissue-marker-clusters, fig.width=12, fig.height=6}
# Original cluster assignments in tissue space
SeuratObject::GetTissueCoordinates(object = seurat) |>
    left_join(seurat[[]], by = join_by(cell == cell_id)) |>
    ggplot(aes(x = x, y = y, color = as.factor(!!sym(params$cluster_name)))) + 
    geom_point(size = 0) ->
    plt_one

# Marker-based sub-clustering in tissue space
SeuratObject::GetTissueCoordinates(object = seurat) |>
    left_join(cluster_seurat[[]], by = join_by(cell == cell_id)) |>
    ggplot(aes(x = x, y = y, color = marker_clusters)) + 
    geom_point(size = 0) ->
    plt_two

# Display side by side
plot_grid(
    plt_one + coord_flip(),
    plt_two + coord_flip()
)
```

# UMAP Analysis - Standard Reduction

These plots use dimensionality reduction based on all genes in the dataset (standard workflow). They help us understand the overall structure and quality of the cluster `r knitr::combine_words(params$coi)` subset.

## Quality Control Metrics

**Quality indicators**:
- **Cell area**:
- **nCount_Xenium**: Total transcript counts; very low values may indicate low-quality cells
- **nFeature_Xenium**: Number of unique genes detected

```{r umap-qc-standard, fig.width=12, fig.height=12}
# Visualize cluster identity and QC metrics in UMAP space

# Cluster assignments
p1 <- DimPlot(
    object = cluster_seurat, 
    group.by = params$cluster_name, 
    reduction = "umap"
)

# Cluster assignments
p2 <- DimPlot(
    object = cluster_seurat, 
    group.by = "seurat_clusters", 
    reduction = "umap"
)

# Cell area
p3 <- FeaturePlot(
    object = cluster_seurat, 
    features = "cell_area", 
    reduction = "umap"
)

# Total counts (library size)
p4 <- FeaturePlot(
    object = cluster_seurat, 
    features = "nCount_Xenium", 
    reduction = "umap"
)

# Number of detected features (genes)
p5 <- FeaturePlot(
    object = cluster_seurat, 
    features = "nFeature_Xenium", 
    reduction = "umap"
)

# Display in grid
plot_grid(plotlist = list(p1, p2, p3, p4, p5), ncol = 2)
```

## Cell Type Scores

```{r umap-cell-type-scores-standard, fig.width=12, fig.height=12}
# Visualize cluster identity and QC metrics in marker-based UMAP space

# Marker-based cluster assignments
p1 <- FeaturePlot(
    object = cluster_seurat, 
    features = "theca_cell_avg", 
    reduction = "umap"
)

# Cell area
p2 <- FeaturePlot(
    object = cluster_seurat, 
    features = "granulosa_cell_avg", 
    reduction = "umap"
)

# Total counts (library size)
p3 <- FeaturePlot(
    object = cluster_seurat, 
    features = "upC1_avg", 
    reduction = "umap"
)

# Number of detected features (genes)
p4 <- FeaturePlot(
    object = cluster_seurat, 
    features = "upC2_avg", 
    reduction = "umap"
)

# Number of detected features (genes)
p5 <- FeaturePlot(
    object = cluster_seurat, 
    features = "upC_diff", 
    reduction = "umap"
)

# Display in grid
plot_grid(plotlist = list(p1, p2, p3, p4, p5), ncol = 2)
```

## Biological Metadata

Do cells from different developmental stages or segmentation quality categories cluster separately?

```{r umap-metadata-standard, fig.width=12, fig.height=6}
# Visualize biological annotations in UMAP space

# Developmental stage labels
plot_grid(
    DimPlot(
        object = cluster_seurat, 
        group.by = "Stage Labels", 
        reduction = "umap"
    ) + 
        theme(legend.position = "bottom"),
    
    # Quality annotations
    DimPlot(
        object = cluster_seurat, 
        group.by = "Quality", 
        reduction = "umap"
    ) + 
        theme(legend.position = "bottom")
)
```

# UMAP Analysis - Marker-based Reduction

These plots use dimensionality reduction based only on curated marker genes. This focused approach might reveal biologically meaningful structure that might be obscured by noise from non-informative genes.

## Quality Control Metrics

**Comparison with standard UMAP**:

```{r umap-qc-markers, fig.width=12, fig.height=12}
# Visualize cluster identity and QC metrics in marker-based UMAP space

# Marker-based cluster assignments
p1 <- DimPlot(
    object = cluster_seurat, 
    group.by = "marker_clusters", 
    reduction = "marker_umap"
)

# Cell area
p2 <- FeaturePlot(
    object = cluster_seurat, 
    features = "cell_area", 
    reduction = "marker_umap"
)

# Total counts (library size)
p3 <- FeaturePlot(
    object = cluster_seurat, 
    features = "nCount_Xenium", 
    reduction = "marker_umap"
)

# Number of detected features (genes)
p4 <- FeaturePlot(
    object = cluster_seurat, 
    features = "nFeature_Xenium", 
    reduction = "marker_umap"
)

# Display in grid
plot_grid(plotlist = list(p1, p2, p3, p4))
```

## Biological Metadata

Examining how developmental stage and quality annotations map onto the marker-based UMAP helps validate whether our marker-driven clustering gives better results.

```{r umap-metadata-markers, fig.width=12, fig.height=6}
# Visualize biological annotations in marker-based UMAP space

# Developmental stage labels
plot_grid(
    DimPlot(
        object = cluster_seurat, 
        group.by = "Stage Labels", 
        reduction = "marker_umap"
    ) + 
        theme(legend.position = "bottom"),
    
    # Quality annotations
    DimPlot(
        object = cluster_seurat, 
        group.by = "Quality", 
        reduction = "marker_umap"
    ) + 
        theme(legend.position = "bottom")
)
```

## Cell Type Scores

```{r umap-cell-type-scores-markers, fig.width=12, fig.height=12}
# Visualize cluster identity and QC metrics in marker-based UMAP space

# Marker-based cluster assignments
p1 <- FeaturePlot(
    object = cluster_seurat, 
    features = "theca_cell_avg", 
    reduction = "marker_umap"
)

# Cell area
p2 <- FeaturePlot(
    object = cluster_seurat, 
    features = "granulosa_cell_avg", 
    reduction = "marker_umap"
)

# Total counts (library size)
p3 <- FeaturePlot(
    object = cluster_seurat, 
    features = "upC1_avg", 
    reduction = "marker_umap"
)

# Number of detected features (genes)
p4 <- FeaturePlot(
    object = cluster_seurat, 
    features = "upC2_avg", 
    reduction = "marker_umap"
)

# Number of detected features (genes)
p5 <- FeaturePlot(
    object = cluster_seurat, 
    features = "upC_diff", 
    reduction = "marker_umap"
)

# Display in grid
plot_grid(plotlist = list(p1, p2, p3, p4, p5), ncol = 2)
```

```{r}
cells_keep <- Embeddings(object = cluster_seurat, "umap")[, 1] < params$filt_umap_1
cluster_cluster_seurat <- cluster_seurat[, cells_keep]
cluster_cluster_seurat
```


```{r}
# Perform standard Seurat workflow on the subset
# This allows us to find sub-structure within the cluster

# Normalize counts
cluster_cluster_seurat <- NormalizeData(object = cluster_cluster_seurat)

# Find highly variable genes
cluster_cluster_seurat <- FindVariableFeatures(object = cluster_cluster_seurat)

# Scale data
cluster_cluster_seurat <- ScaleData(object = cluster_cluster_seurat)

# Run PCA on all genes
cluster_cluster_seurat <- RunPCA(object = cluster_cluster_seurat)

# # Run PCA using only marker genes
# cluster_cluster_seurat <- RunPCA(
#     object = cluster_cluster_seurat, 
#     features = unique(marker_dat$Gene), 
#     reduction.name = "pca_markers"
# )

# Find neighbors using standard PCA
cluster_cluster_seurat <- FindNeighbors(object = cluster_cluster_seurat, dims = 1:30)

# # Find neighbors using marker-based PCA
# cluster_cluster_seurat <- FindNeighbors(
#     object = cluster_cluster_seurat, 
#     dims = 1:30, 
#     reduction = "pca_markers", 
#     graph.name = "marker_graph"
# )

# Cluster using standard graph
cluster_cluster_seurat <- FindClusters(object = cluster_cluster_seurat)

# # Cluster using marker-based graph
# cluster_cluster_seurat <- FindClusters(
#     object = cluster_cluster_seurat, 
#     graph.name = "marker_graph", 
#     cluster.name = "marker_clusters"
# )

# Generate UMAP from standard PCA
cluster_cluster_seurat <- RunUMAP(object = cluster_cluster_seurat, dims = 1:30)

# # Generate UMAP from marker-based PCA
# cluster_cluster_seurat <- RunUMAP(
#     object = cluster_cluster_seurat, 
#     dims = 1:30, 
#     reduction = "pca_markers", 
#     reduction.name = "marker_umap"
# )
```



```{r}
# Visualize cluster identity and QC metrics in UMAP space

# Cluster assignments
p1 <- DimPlot(
    object = cluster_cluster_seurat, 
    group.by = params$cluster_name, 
    reduction = "umap"
)

# Cluster assignments
p2 <- DimPlot(
    object = cluster_cluster_seurat, 
    group.by = "seurat_clusters", 
    reduction = "umap"
)

# Cell area
p3 <- FeaturePlot(
    object = cluster_cluster_seurat, 
    features = "cell_area", 
    reduction = "umap"
)

# Total counts (library size)
p4 <- FeaturePlot(
    object = cluster_cluster_seurat, 
    features = "nCount_Xenium", 
    reduction = "umap"
)

# Number of detected features (genes)
p5 <- FeaturePlot(
    object = cluster_cluster_seurat, 
    features = "nFeature_Xenium", 
    reduction = "umap"
)

# Display in grid
plot_grid(plotlist = list(p1, p2, p3, p4, p5), ncol = 2)
```