---
title: "Point Pattern Analysis – Univariate Methods for Imaging-Based Data"
author: "Adapted from Robinson Lab PASTA (robinsonlabuzh.github.io/pasta)"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    self-contained: true
params:
  # Path to your SpatialExperiment .rds file
  spe_path: "../data/spe.rds"
  # Column in colData() that identifies images/sections (e.g. sample_id, z_stack)
  image_col: "sample_id"
  # Which image IDs to include in the analysis (list of strings)
  image_ids: ["-0.09", "0.01", "0.21"]
  # Column in colData() containing cell type / categorical mark labels
  mark_col: "cluster_id"
  # Cell type of interest for univariate analysis
  celltype: "OD Mature"
  # Additional grouping variables used in calcMetricPerFov (besides image_col)
  by_vars: ["Animal_ID", "sample_id"]
  # Genes to use for continuous-mark analysis (must be in rownames of assay)
  genes: ["Slc18a2", "Esr1", "Pgr"]
  # Image ID to use for single-image analyses (LISA, fPCA, continuous marks)
  focal_image: "0.01"
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false
library(SpatialExperiment)
library(spatialFDA)
library(spatstat)
library(sosta)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(reshape2)
library(magrittr)
library(refund)

# ── Helper: convert SpatialExperiment to spatstat ppp ───────────────────────
# This mimics the .ppp() utility from utils.R in the original PASTA repo.
# Adjust if your coordinate column names differ (default: x, y).
spe_to_ppp <- function(spe_obj, mark_col = "cluster_id") {
  coords <- spatialCoords(spe_obj)
  x <- coords[, 1]
  y <- coords[, 2]
  marks_vec <- factor(colData(spe_obj)[[mark_col]])
  ppp(x, y, range(x), range(y), marks = marks_vec)
}
```

## Overview

This notebook applies **univariate point pattern analysis** to imaging-based spatial omics data.  
It is a generalized version of the [PASTA vignette](https://robinsonlabuzh.github.io/pasta/01-imaging-univar-ppSOD.html) by the Robinson Lab, UZH.

**What this notebook covers:**

- Observation window estimation (Ripley–Rasson and density-based via `sosta`)
- Correlation metrics: Ripley's *K*, *L*-function, pair correlation function (homogeneous, inhomogeneous, locally scaled)
- Local indicators of spatial association (LISA *L*-function) and functional PCA
- Spacing metrics: *G*, *F*, *J* functions (homogeneous and inhomogeneous), nearest-neighbour orientation
- Continuous mark summary functions: mark correlation and mark-weighted *K*

**Key parameters** (set in the YAML `params` block above):

| Parameter | Current value | Description |
|-----------|---------------|-------------|
| `spe_path` | `{r params$spe_path}` | Path to `SpatialExperiment` RDS |
| `image_col` | `{r params$image_col}` | Column identifying images |
| `image_ids` | `{r paste(params$image_ids, collapse=", ")}` | Images to analyse |
| `mark_col` | `{r params$mark_col}` | Categorical mark column |
| `celltype` | `{r params$celltype}` | Cell type of interest |
| `focal_image` | `{r params$focal_image}` | Image for single-image analyses |
| `genes` | `{r paste(params$genes, collapse=", ")}` | Genes for continuous marks |

---

# Part 1 – Categorical Marks

## 1.1 Load data and build point patterns

```{r load-data}
spe <- readRDS(params$spe_path)

# Select images of interest
spe_sub <- spe[, colData(spe)[[params$image_col]] %in% params$image_ids]

# Build a named list of ppp objects (one per image), marks = cell type label
build_ppp_list <- function(image_id, spe_obj, image_col, mark_col) {
  sub <- spe_obj[, colData(spe_obj)[[image_col]] == image_id]
  spe_to_ppp(sub, mark_col = mark_col)
}

pp_raw <- lapply(params$image_ids, build_ppp_list,
                 spe_obj = spe_sub,
                 image_col = params$image_col,
                 mark_col = params$mark_col)
names(pp_raw) <- params$image_ids
```

## 1.2 Observation Windows

### Ripley–Rasson convex window estimate

Most often, only a subset of the full point process is observed ("window sampling"). The Ripley–Rasson estimator provides a convex approximation of the observation window.

```{r ripras-windows}
set_ripras <- function(pp_obj) {
  Window(pp_obj) <- ripras(pp_obj)
  pp_obj
}

# Apply Ripley-Rasson window, then split by mark (cell type)
pp_ripras <- lapply(pp_raw, set_ripras)
pp_ls     <- lapply(pp_ripras, split)   # list of lists: image -> celltype -> ppp
```

### Density-based window estimate (`sosta`)

An alternative is to estimate the window from the spatial intensity of cells using a kernel density threshold.

```{r sosta-windows, fig.width=14, fig.height=5}
n_params <- estimateReconstructionParametersSPE(
  spe_sub,
  marks     = params$mark_col,
  imageCol  = params$image_col,
  markSelect = NULL,
  plotHist  = FALSE
)

thres_spe <- mean(n_params$thres)
bndw_spe  <- mean(n_params$bndw)

all_structs <- reconstructShapeDensitySPE(
  spe_sub,
  marks     = params$mark_col,
  imageCol  = params$image_col,
  markSelect = NULL,
  bndw      = bndw_spe,
  thres     = thres_spe,
  nCores    = 1
)

# Plot estimated windows overlaid on cell positions
p_windows <- lapply(params$image_ids, function(img) {
  sub_img    <- spe_sub[, colData(spe_sub)[[params$image_col]] == img]
  structs_img <- subset(all_structs, get(params$image_col) == img)

  cbind(as.data.frame(colData(sub_img)), spatialCoords(sub_img)) |>
    ggplot(aes(x = x, y = y)) +
    geom_point(size = 0.25) +
    geom_sf(data = structs_img, fill = NA, color = "red",
            inherit.aes = FALSE, linewidth = 0.8) +
    facet_grid(reformulate(params$image_col)) +
    theme_light() +
    ggtitle(img)
})

wrap_plots(p_windows)
```

---

## 1.3 Univariate viewpoint

We compare the spatial distribution of **`r params$celltype`** across images  
`r paste(params$image_ids, collapse = ", ")`, treating each image as an independent realisation.

---

## 1.4 Correlation Functions

### Ripley's *K*, *L*-function, and Pair Correlation Function (homogeneous)

The *K*-function counts the average number of neighbours within radius *r* per point, normalised by intensity. The *L*-function is a variance-stabilising transformation. The pair correlation function (pcf) is the derivative of *K* and shows contributions at exactly distance *r*.

```{r homo-correlation, fig.width=16, fig.height=6}
res_homo <- lapply(c("Kest", "Lest", "pcf"), function(fun) {
  res <- calcMetricPerFov(
    spe,
    params$celltype,
    subsetby = params$image_col,
    fun      = fun,
    marks    = params$mark_col,
    rSeq     = NULL,
    by       = params$by_vars
  )
  subset(res, get(params$image_col) %in% params$image_ids)
})

p_homo_ls <- lapply(res_homo, function(res) {
  plotMetricPerFov(res, theo = TRUE, correction = "iso",
                   x = "r", imageId = params$image_col,
                   legend.position = "right")
})

p_homo <- wrap_plots(p_homo_ls, guides = "collect")
p_homo
```

### Inhomogeneous *K*, *L*, and pcf

When spatial intensity varies (inhomogeneity), the standard *K*-function is biased. The inhomogeneous versions correct for locally estimated intensity via kernel smoothing.

```{r inhomo-correlation, fig.width=16, fig.height=6}
res_inhomo <- lapply(c("Kinhom", "Linhom", "pcfinhom"), function(fun) {
  res <- calcMetricPerFov(
    spe,
    params$celltype,
    subsetby  = params$image_col,
    fun       = fun,
    marks     = params$mark_col,
    rSeq      = NULL,
    by        = params$by_vars,
    correction = "isotropic"
  )
  subset(res, get(params$image_col) %in% params$image_ids)
})

p_inhomo_ls <- lapply(res_inhomo, function(res) {
  plotMetricPerFov(res, correction = "iso", theo = TRUE,
                   x = "r", imageId = params$image_col,
                   legend.position = "right")
})

p_inhomo <- wrap_plots(p_inhomo_ls, guides = "collect")
p_inhomo
```

### Locally-Scaled *K* and *L*

Local scaling further accounts for cases where the point process in different sub-regions is a scaled version of a common template process.

```{r scaled-correlation, fig.width=16, fig.height=6}
res_scaled <- lapply(c("Kscaled", "Lscaled"), function(fun) {
  res <- calcMetricPerFov(
    spe,
    params$celltype,
    subsetby = params$image_col,
    fun      = fun,
    marks    = params$mark_col,
    rSeq     = NULL,
    by       = params$by_vars
  )
  subset(res, get(params$image_col) %in% params$image_ids)
})

p_scaled_ls <- lapply(res_scaled, function(res) {
  plotMetricPerFov(res, correction = "iso", theo = TRUE,
                   x = "r", imageId = params$image_col,
                   legend.position = "right")
})

p_scaled <- wrap_plots(p_scaled_ls, guides = "collect")
p_scaled
```

### All correlation variants combined

```{r all-correlation, fig.width=16, fig.height=14}
wrap_plots(list(p_homo, p_inhomo, p_scaled),
           nrow = 3, guides = "collect") +
  plot_annotation(tag_levels = "A")
```

---

### Local Indicators of Spatial Association (LISA *L*-function)

LISA computes the *L*-function contribution for each individual point, allowing detection of heterogeneous spatial structure within a single image.

```{r lisa, fig.width=10, fig.height=5}
focal_pp <- pp_ls[[params$focal_image]][[params$celltype]]

L_lisa <- localL(focal_pp)
df_lisa <- as.data.frame(L_lisa)
dfm     <- reshape2::melt(df_lisa, "r")

# Pick a reference radius near the middle of the range for colouring
ref_r <- median(df_lisa$r)
get_sel <- dfm |>
  filter(abs(r - ref_r) == min(abs(r - ref_r)), variable != "theo") |>
  mutate(sel = value) |>
  select(variable, sel)

dfm <- dfm |> left_join(get_sel, by = "variable")

p_lisa <- ggplot(dfm, aes(x = r, y = value, group = variable, colour = sel)) +
  geom_line() +
  scale_color_continuous(type = "viridis") +
  geom_vline(xintercept = ref_r, linetype = "dashed") +
  theme_light() +
  ggtitle(paste0("LISA L-curves – image ", params$focal_image))

ppdf <- as.data.frame(pp_ripras[[params$focal_image]]) |>
  filter(marks == params$celltype)
ppdf$sel <- get_sel$sel

q_lisa <- ggplot(ppdf, aes(x = x, y = y, colour = sel)) +
  geom_point(size = 0.75) +
  scale_color_continuous(type = "viridis") +
  theme_light() +
  ggtitle(paste0("Points coloured by LISA value at r ~ ", round(ref_r)))

p_lisa | q_lisa
```

### Functional PCA of LISA Curves

Functional PCA extracts the dominant modes of variation across the *n* individual LISA curves.

```{r fpca, fig.width=10, fig.height=5}
df_fdob <- asinh(df_lisa |> as.matrix() / 50) |> as.data.frame()

mat <- df_fdob |>
  select(!c(r, theo)) |>
  t()

dat <- data.frame(ID = rownames(mat))
dat$Y   <- mat
dat$sel <- get_sel$sel

res_fpca <- functionalPCA(dat = dat, r = unique(df_fdob$r))

scores_df <- as.data.frame(res_fpca$scores)

p_biplot <- ggplot(scores_df, aes(scores_df[, 1], scores_df[, 2],
                                   colour = dat$sel)) +
  geom_point() +
  coord_equal() +
  theme_light() +
  scale_color_continuous(type = "viridis") +
  xlab("PC1") + ylab("PC2") +
  ggtitle("fPCA biplot of LISA curves")

p_biplot
```

```{r fpca-maps, fig.width=10, fig.height=5}
p_pc1 <- ggplot(ppdf, aes(x = x, y = y, colour = res_fpca$scores[, 1])) +
  scale_color_continuous(type = "viridis", name = "PC1 loading") +
  geom_point(size = 0.75) + theme_light() + ggtitle("PC1 loadings on tissue")

p_pc2 <- ggplot(ppdf, aes(x = x, y = y, colour = res_fpca$scores[, 2])) +
  scale_color_continuous(type = "viridis", name = "PC2 loading") +
  geom_point(size = 0.75) + theme_light() + ggtitle("PC2 loadings on tissue")

p_pc1 | p_pc2
```

---

## 1.5 Spacing Functions

### Mean Nearest-Neighbour Distance (MNND)

```{r mnnd, fig.width=8, fig.height=5}
nndistance <- function(pp_obj, nk) {
  xy <- cbind(pp_obj$x, pp_obj$y)
  rowMeans(nndist(xy, k = nk))
}

# Use up to 3 cell types from the focal image for illustration; adjust as needed
available_types <- names(pp_ls[[params$focal_image]])
celltypes_plot  <- head(available_types, 3)

res_nnd <- lapply(celltypes_plot, function(ct) {
  vals <- nndistance(pp_ls[[params$focal_image]][[ct]], seq_len(15))
  data.frame(metric = vals, type = ct)
})

res_nnd_df <- do.call(rbind, res_nnd)

ggplot(res_nnd_df, aes(x = metric, col = type)) +
  geom_density(linewidth = 1) +
  scale_x_sqrt() +
  theme_light() +
  ggtitle("Sqrt of Mean Nearest-Neighbour Distance") +
  xlab("sqrt(MNND)")
```

### *G*, *F*, and *J* Functions (homogeneous)

The *G*-function is the nearest-neighbour distance distribution; *F* is the empty-space function; *J* combines both (J = 1 for CSR, J < 1 = clustering, J > 1 = regularity).

```{r GFJ-homo, fig.width=16, fig.height=6}
res_gfj <- lapply(c("Gest", "Fest", "Jest"), function(fun) {
  res <- calcMetricPerFov(
    spe,
    params$celltype,
    subsetby = params$image_col,
    fun      = fun,
    marks    = params$mark_col,
    rSeq     = NULL,
    by       = params$by_vars
  )
  subset(res, get(params$image_col) %in% params$image_ids)
})

p_gfj_ls <- lapply(res_gfj, function(res) {
  plotMetricPerFov(res, correction = "rs", theo = TRUE,
                   x = "r", imageId = params$image_col,
                   legend.position = "right")
})

wrap_plots(p_gfj_ls, guides = "collect")
```

### *G*, *F*, and *J* Functions (inhomogeneous)

```{r GFJ-inhomo, fig.width=16, fig.height=6}
res_gfj_inhom <- lapply(c("Ginhom", "Finhom", "Jinhom"), function(fun) {
  res <- calcMetricPerFov(
    spe,
    params$celltype,
    subsetby = params$image_col,
    fun      = fun,
    marks    = params$mark_col,
    rSeq     = NULL,
    by       = params$by_vars
  )
  subset(res, get(params$image_col) %in% params$image_ids)
})

p_gfj_inhom_ls <- lapply(res_gfj_inhom, function(res) {
  plotMetricPerFov(res, correction = "bord", theo = TRUE,
                   x = "r", imageId = params$image_col,
                   legend.position = "right")
})

wrap_plots(p_gfj_inhom_ls, guides = "collect")
```

### Nearest-Neighbour Orientation

Orientation of *k*-th nearest neighbours reveals isotropy/anisotropy in the point pattern.

```{r nnorient, fig.width=8, fig.height=5}
res_nnorient <- calcMetricPerFov(
  spe,
  params$celltype,
  subsetby = params$image_col,
  fun      = "nnorient",
  marks    = params$mark_col,
  rSeq     = NULL,
  by       = params$by_vars
)
res_nnorient <- subset(res_nnorient, get(params$image_col) %in% params$image_ids)

plotMetricPerFov(res_nnorient, correction = "bordm", theo = TRUE,
                 x = "phi", imageId = params$image_col,
                 legend.position = "right")
```

```{r rose-diagrams, fig.width=8, fig.height=6}
# Rose diagrams per image
lapply(seq_along(pp_ripras), function(i) {
  rose(nnorient(pp_ripras[[i]]), main = names(pp_ripras)[i])
})
```

---

# Part 2 – Continuous Marks

## 2.1 Setup

We attach gene expression values as continuous marks to the point pattern of the focal image (`r params$focal_image`).

```{r continuous-setup, fig.width=12, fig.height=5}
sub_focal <- spe[, colData(spe)[[params$image_col]] == params$focal_image]
pp_focal  <- pp_ripras[[params$focal_image]]

# Attach gene expression as marks
gex <- assay(sub_focal)[params$genes, ] |>
  t() |>
  as.matrix() |>
  as.data.frame()
rownames(gex) <- NULL
marks(pp_focal) <- gex

# Visualise spatial gene expression
pp_focal |>
  as.data.frame() |>
  tidyr::pivot_longer(cols = seq_along(params$genes) + 2) |>
  ggplot(aes(x, y, colour = log(value + 1))) +
  geom_point(size = 0.5) +
  facet_wrap(~name) +
  coord_equal() +
  scale_color_continuous(type = "viridis") +
  theme_light() +
  ggtitle(paste0("Log gene expression – image ", params$focal_image))
```

```{r pairs-plot, fig.width=8, fig.height=8}
pairs(as.data.frame(pp_focal), panel = panel.smooth, pch = ".")
```

```{r nnmark-plot, fig.width=8, fig.height=8}
plot(nnmark(pp_focal))
```

---

## 2.2 Mark Correlation Function

The mark correlation function measures mark dependence between pairs of points at distance *r*. Values > 1 indicate positive mark correlation; < 1 indicates negative correlation; 1 = random labelling.

```{r markcorr, fig.width=10, fig.height=5}
res_mcorr <- calcMetricPerFov(
  sub_focal,
  params$celltype,
  subsetby   = params$genes,
  fun        = "markcorr",
  marks      = params$genes,
  rSeq       = NULL,
  by         = params$by_vars,
  continuous = TRUE
)

plotMetricPerFov(res_mcorr, correction = "iso", theo = TRUE,
                 x = "r", imageId = "gene",
                 legend.position = "right")
```

---

## 2.3 Mark-Weighted *K*-Function

The mark-weighted *K*-function generalises Ripley's *K* by weighting each pair contribution by the product of their marks. Comparing this to the unweighted *L*-function reveals whether high-expression cells are more spatially clustered than expected.

```{r kmark, fig.width=10, fig.height=5}
res_kmark <- calcMetricPerFov(
  sub_focal,
  params$celltype,
  subsetby   = params$genes,
  fun        = "Kmark",
  marks      = params$genes,
  rSeq       = NULL,
  by         = params$by_vars,
  continuous = TRUE
)

plotMetricPerFov(res_kmark, correction = "iso", theo = TRUE,
                 x = "r", imageId = "gene",
                 legend.position = "right")
```

### Mark-weighted *L* vs. unmarked *L* for a single gene

Comparing the mark-weighted *L* with the standard *L* isolates the contribution of the continuous mark to spatial clustering.

```{r kmark-vs-L, fig.width=7, fig.height=5}
focal_gene  <- params$genes[1]   # change index to inspect a different gene
pp_gene     <- subset(pp_focal, select = focal_gene)

L_marked  <- Kmark(pp_gene, function(m1, m2) m1 * m2, returnL = TRUE)
L_plain   <- Lest(pp_gene, nlarge = 7000)

plot(eval.fv(L_marked - L_plain),
     main = paste0("Mark-weighted L minus plain L (", focal_gene, ")"))
```

---

# Summary

| Analysis | Functions used |
|----------|----------------|
| Observation windows | `ripras`, `reconstructShapeDensitySPE` |
| Homogeneous correlation | `Kest`, `Lest`, `pcf` |
| Inhomogeneous correlation | `Kinhom`, `Linhom`, `pcfinhom` |
| Locally-scaled | `Kscaled`, `Lscaled` |
| LISA + fPCA | `localL`, `functionalPCA` |
| Spacing | `Gest`/`Ginhom`, `Fest`/`Finhom`, `Jest`/`Jinhom` |
| NN orientation | `nnorient`, `rose` |
| Continuous marks | `markcorr`, `Kmark` |

---

# Session Info

```{r session-info}
sessionInfo()
```
